---
title: "Reproducible Research: Peer Assessment 2 - Storm Data Analysis"
output: 
  html_document:
    keep_md: true
---

## Introduction

#This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Synopsis

#In this project we try to explore the NOAA data storm database (1950 - 2011). Our prime objective is to find out what events are most significant in terms of fatalities, injuries and economic damage (property and crop). We'll focus our analysis to answer below two questions:
#Q1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
#Q2. Across the United States, which types of events have the greatest economic consequences?



##  Reading Dataset

```{r}

library(ggplot2)
library(knitr)
library(dplyr)

#getwd()
#dir()


storm<- read.csv2("repdata_data_StormData.csv",  header = TRUE, sep = ",")

#head(storm)
summary(storm)

```


##  Data Processing

```{r}

#Trim the dataset to keep only relevant columns

storm_sub <- storm[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
    "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

head(storm_sub)

# Create subset for Question 1 & 2

# Select data for Fatalities and injuries from subset (Q1)

harm <- subset(storm_sub, !storm_sub$FATALITIES == 0 & !storm_sub$INJURIES == 
    0, select = c(EVTYPE, FATALITIES, INJURIES))

harm$FATALITIES <- as.numeric(as.character(harm$FATALITIES))
harm$INJURIES <- as.numeric(as.character(harm$INJURIES))

head(harm)

# Select data for various damages from subset (Q2)

Dam <- subset(storm_sub, !storm_sub$PROPDMG == 0 & !storm_sub$CROPDMG == 
    0, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

head(Dam)

```

## Q1 - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}

# Create separate data set for Fatalities

fatal <- harm %>%
          group_by(EVTYPE) %>%
          summarise_each(funs(sum))%>% 
          arrange(desc(FATALITIES))
         
#head(fatal)

# Take top 10 events for Fatalities and corresponding Injuries

fatal <- fatal[1:10,]

# Create separate data set for Injury

Inj <- harm %>%
          group_by(EVTYPE) %>%
          summarise_each(funs(sum))%>% 
          arrange(desc(INJURIES))
         
#head(Inj)

# Take top 10 events for Injuries

Inj <- Inj[1:10,]
#head(Inj)

# Plot Fatalities and store at fatal_plot
library(ggplot2)

fatal_plot <- ggplot() + geom_bar(data = fatal, aes(x = EVTYPE, 
    y =FATALITIES, fill = interaction(FATALITIES, EVTYPE)), stat = "identity", 
    show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    xlab("Harmful Events") + ylab("FATALITIES") + ggtitle("Top 10 weather events causing FATALITIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Plot Fatalities and store at Inj_plot
Inj_plot <- ggplot() + geom_bar(data = Inj, aes(x = EVTYPE, 
    y =INJURIES, fill = interaction(INJURIES, EVTYPE)), stat = "identity", 
    show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    xlab("Harmful Events") + ylab("INJURIES") + ggtitle("Top 10 weather events causing INJURIES") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

#Show both plots for Fatalities and Injuries

require("gridExtra")
grid.arrange(fatal_plot, Inj_plot, ncol = 2)

```
## Assessment for Q1: Tornados are the most harmful in terms of Fatality and Injury based on given data

## Q2 - Across the United States, which types of events have the greatest economic consequences?

```{r}
# select required entries for economy damage
Dam <- subset(Dam, Dam$PROPDMGEXP == "K" | Dam$PROPDMGEXP == 
    "k" | Dam$PROPDMGEXP == "M" | Dam$PROPDMGEXP == "m" | 
    Dam$PROPDMGEXP == "B" | Dam$PROPDMGEXP == "b")

Dam <- subset(Dam, Dam$CROPDMGEXP == "K" | Dam$CROPDMGEXP == 
    "k" | Dam$CROPDMGEXP == "M" | Dam$CROPDMGEXP == "m" | 
    Dam$CROPDMGEXP == "B" | Dam$CROPDMGEXP == "b")

# Convert economic values to number
Dam$PROPDMGEXP <- gsub("m", 1e+06, Dam$PROPDMGEXP, ignore.case = TRUE)
Dam$PROPDMGEXP <- gsub("k", 1000, Dam$PROPDMGEXP, ignore.case = TRUE)
Dam$PROPDMGEXP <- gsub("b", 1e+09, Dam$PROPDMGEXP, ignore.case = TRUE)
Dam$PROPDMGEXP <- as.numeric(Dam$PROPDMGEXP)
Dam$CROPDMGEXP <- gsub("m", 1e+06, Dam$CROPDMGEXP, ignore.case = TRUE)
Dam$CROPDMGEXP <- gsub("k", 1000, Dam$CROPDMGEXP, ignore.case = TRUE)
Dam$CROPDMGEXP <- gsub("b", 1e+09, Dam$CROPDMGEXP, ignore.case = TRUE)
Dam$CROPDMGEXP <- as.numeric(Dam$CROPDMGEXP)
Dam$PROPDMGEXP <- as.numeric(Dam$PROPDMGEXP)
Dam$CROPDMG <- as.numeric(Dam$CROPDMG)
Dam$PROPDMG <- as.numeric(Dam$PROPDMG)

# then sum the damages by each event type
Dam$TOTALDMG <- (Dam$CROPDMG * Dam$CROPDMGEXP) + 
    (Dam$PROPDMG * Dam$PROPDMGEXP)

Dam <- aggregate(Dam$TOTALDMG, by = list(Dam$EVTYPE), 
    FUN = sum)

colnames(Dam) <- c("EVTYPE", "TOTALDMG")

# Rank the event type by highest damage cost and take top 5 columns
Dam <- Dam[order(Dam$TOTALDMG, decreasing = TRUE), 
    ]
Dam <- Dam[1:5, ]

# Now plot the graph
ggplot() + geom_bar(data = Dam, aes(x = EVTYPE, y = TOTALDMG, fill = interaction(TOTALDMG, 
    EVTYPE)), stat = "identity", show.legend = F) + theme(axis.text.x = element_text(angle = 30, 
    hjust = 1)) + xlab("Event Type") + ylab("Total Damage")


```
## Q2 Assessment: Flood brings most economic damage as per given data.

## Result

```{r}

# As per the analysis of given data, we can infer that:

#1. Tornados are most harmful event in terms of fatality and Injuries.
#2. Flood is most harmful even in terms of economic damage.

```